Exploring United States Housing Affordability

### 2016-08-10 paper/Paper.Rmd
final changes hopefully

The dataset was taken from the American Housing Survey (AHS), which derives sets of data from 1985 and on. The data set includes categorizing each housing unit by factors such as affordability, income through fair market rent, burden, and the adjusted median income. Certain factors, such as Poverty Income, is based on the Census Bureau's official thresholds.

@@

```{r, echo = FALSE, fig.cap="Relationship between Household Income and Burden, fig.height = 1"}
###### Household Income
  ####overall graph
clean_years %>% 
  select(zinc2, 
         year, 
         region, 
         fmtburden)%>% 
  filter(fmtburden != ".", 
         fmtburden != "No Income") %>% 
  mutate(fmtburden = str_replace(fmtburden, 
                                 "Greater than 50%", 
                                 "50% or More")) %>% 
  group_by_("year", "fmtburden") %>%  
  summarise_(new_y = as.formula(
  paste0("~ median(","zinc2", ",na.rm=T)"))) %>% 
  ggplot(aes_string(x = "year", 
                    y = "new_y", 
                    color = "fmtburden")) + 
  geom_line(size = 1) + 
  labs(y = "Household Income", 
       x = "Year", 
       color = "Burden") +
  scale_y_continuous(limits = c(0,NA), 
                     breaks=c(0,20000,40000,60000),
                     labels=function(x){paste0("$",x/1000, "K")}) +
  scale_colour_brewer(palette = "Dark2") +
  theme_minimal(base_family = "serif")

@@

  As pictured in figure 5, the relationship between household income and burden is seen as a positive correlation. From the graph, you can see that higher income household have less burden to deal with, which makes sense because those with more money are able to pay off monthly costs more easily. Those with 50% more burden are near the poverty line. Figure 5 also shows a spike in households that have burden. Households with higher incomes started to pay more burdens, which can be attributed to 2007 financial housing market crisis.

@@

cleanyearsmonthly %>%  
  group_by_("year", "fmtburden") %>%  
  summarise_(new_y = as.formula(
    paste0("~ median(","zsmhc", ", na.rm=T)"))) %>%

@@

At a certain point, regardless of income, housing costs stay the same because housing prices have increased

The no income line continues to have monthly housing costs due to people who lost their jobs, unemployment, trust fund recipients using their money, and people living off savings

  Figure 6 offers a more in-depth look at the relationship between housing costs and burden. It's interesting to see that all three burden lines are close to one another, which illustrates the principle that those with more money aren't buying more expensive houses. From figure 5 we saw that those with higher incomes have less burden, and those with lower incomes have higher burden. Each level of burden has similar housing costs, therefore reinforcing the idea that people are buying similar priced housing among the lower and middle classes. Again, a spike around 2010 is prevalent, which can be caused by the 2007 housing crisis.

@@

### faceted by region
  cleanyearsmonthly %>%  
  group_by_("year","region", "fmtburden") %>%  
  summarise_(new_y = as.formula(
    paste0("~ median(","zsmhc", ",na.rm=T)"))) %>% 
  ggplot(aes_string(x = "year", y = "new_y", color = "fmtburden")) + 
  geom_line(size = 1) + 
  facet_wrap(~region, labeller = as_labeller(regionNames)) + 
  labs(y = "Monthly Housing Costs", x = "Year", color = "Burden") +
  scale_y_continuous(labels=function(x){paste0("$", x)}) +
  scale_colour_brewer(palette = "Dark2") +
  theme_minimal(base_family = "serif")

@@

  From figure 7 you can see that the Northeast and West have higher rates of burden at higher income levels, therefore illustrating how housing affordability in these regions are less than that of the Midwest and South.

```{r, fig.width=12, fig.height=6,echo = F, message=F,error=F, warning=FALSE, fig.cap = "Income compared to area's median grouped by house type"}
clean_years %>% 
  filter(fmtstructuretype != ".") %>% 
      # Use NSE
      group_by_("year", "fmtstructuretype") %>%  
      # Take the median instead of mean to avoid extreme observations skewing the 
      # result.
      summarise_(new_y= as.formula(paste0("~ median(","increlamipct", ",na.rm=T)"))) %>%  
      # Use NSE
      ggplot(aes_string(x = "year",   
                        y = "new_y",   
                        colour = "fmtstructuretype")) +  
  geom_line(size = 1) +
  scale_y_continuous(labels = function(x){paste0(x, "%")}) +
  labs(colour = "Structure Type",
       y = "Income Percentage of Area Median Income",
       x = "Year") +
  scale_colour_brewer(palette = "Dark2") +
  theme_minimal(base_family = "serif")

```
\begin{center}
Figure 8: Income compared to area's median grouped by house type
\end{center}

@@

```{r, fig.width=12, fig.height=6, echo = F, message=F,error=F, warning=FALSE, fig.cap="Burden based on Median housing cost with assistance"}
assisted = clean_years %>% filter(!is.na(assisted)) %>% 
      # Use NSE
      group_by_("assisted", "fmtcostmedrelamicat") %>%  
      # Take the median instead of mean to avoid extreme observations skewing the 
      # result.
      summarise_(new_y= as.formula(paste0("~ median(","burden", ",na.rm=T)"))) %>% 
      ungroup %>% 
      mutate(assisted = as.factor(assisted))
ggplot(assisted, aes(y = new_y,   
                  x = fmtcostmedrelamicat,
                  fill = assisted, group = assisted)) +  
  geom_bar(
           stat="identity") + 
  labs(y="Burden",
       x = "",
       fill="Assistance") + 
  scale_y_continuous(limits=c(0,1), labels=function(x){paste0(x*100, "%")}) +
  scale_x_discrete(labels = function(x){str_replace_all(str_replace_all(x, "^[0-9] (.*)","\\1"), "LTE (.*)","<= \\1")}) +
  scale_fill_brewer("Assistance", 
                    labels = c("Without Assistance", 
                               "With Assistance"),
                    palette = "Set2")  +
  coord_flip() +
  theme_minimal(base_family = "serif")
```
\begin{center}
Figure 9: Burden based on Median housing cost with assistance
\end{center}

@@

Ultimately, our findings point towards

@@

2.

>>>>

The dataset was taken from the American Housing Survey (AHS), which derives sets of data from 1985 and on. The data set includes categorizing each housing unit by factors such as affordability, income through fair market rent, burden, and the adjusted median income. Certain factors, such as poverty income, is based on the Census Bureau's official thresholds.

@@

```{r, echo = FALSE, fig.cap="Relationship between Household Income and Burden, fig.height = 3"}
###### Household Income
  ####overall graph
clean_years %>% 
  select(zinc2, 
         year, 
         region, 
         fmtburden)%>% 
  filter(fmtburden != ".", 
         fmtburden != "No Income") %>% 
  mutate(fmtburden = str_replace(fmtburden, 
                                 "Greater than 50%", 
                                 "50% or More")) %>% 
  group_by_("year", "fmtburden") %>%  
  summarise_(new_y = as.formula(
  paste0("~ median(","zinc2", ",na.rm=T)"))) %>% 
  ggplot(aes_string(x = "year", 
                    y = "new_y", 
                    color = "fmtburden")) + 
  geom_line(size = 1) + 
  labs(y = "Household Income", 
       x = "Year", 
       color = "Burden") +
  scale_y_continuous(limits = c(0,NA), 
                     breaks=c(0,20000,40000,60000),
                     labels=function(x){paste0("$",x/1000, "K")}) +
  scale_colour_brewer(palette = "Dark2") +
  theme_minimal(base_family = "serif")

@@

  As pictured in Figure 5, the relationship between household income and burden is seen as a positive correlation. The relatively higher household income brackets are shown to have lower burdens than the lower household income brackets. Because burden is a proportion of housing cost over income, it is reasonable to conclude that higher income households will have a smaller burden due to their larger incomes. The higher income households are able to pay off their monthly costs more easily. Although the federal poverty line differs between different sized households, the average poverty threshold for families of 3 or 4 persons is ranged from $20k to $25K. ^2^ From Figure 5, it is easily seen that those with 50% or More burden are near and/or underneath the poverty threshold. At the onset of the 21st century to 2010, there is a spike in 2007/2008 for the range of household income with the differing burden brackets. This spike can be attributed to the 2007 financial housing crisis, which ultimately affected the  burden levels of the housing sector.

@@

cleanyearsmonthly %>%  
  group_by_("year", "fmtburden") %>%  
  summarise_(new_y = as.formula(
    paste0("~ median(",
           "zsmhc", 
           ", na.rm=T)"))) %>%

@@

  Figure 6 offers a more in-depth look at the relationship between housing costs and burden. It's interesting to see that all three burden lines are close to one another, which illustrates the principle that those with more money aren't buying more expensive houses. From Figure 5, we saw that those with higher incomes have less burden, and those with lower incomes have higher burden. At a certain point, regardless of income, housing costs stay the same. For the "No Income" bracket, there are still monthly costs present. To account for this anomly, we assum that people who lost their jobs could still be paying these costs with zero income. Trust fund recipients and people living off savings could also be paying off these costs. Again, a spike around 2010 is prevalent, which can be explained by the 2007 housing crisis.

@@

### faceted by region
  cleanyearsmonthly %>%  
  group_by_("year",
            "region", 
            "fmtburden") %>%  
  summarise_(new_y = as.formula(
    paste0("~ median(",
           "zsmhc", 
           ",na.rm=T)"))) %>% 
  ggplot(aes_string(x = "year", 
                    y = "new_y", 
                    color = "fmtburden")) + 
  geom_line(size = 1) + 
  facet_wrap(~region, 
             labeller = as_labeller(regionNames)) + 
  labs(y = "Monthly Housing Costs", 
       x = "Year", 
       color = "Burden") +
  scale_y_continuous(labels = function(x){paste0("$", x)}) +
  scale_colour_brewer(palette = "Dark2") +
  theme_minimal(base_family = "serif")

@@

  Figure 7 illustrates the monthly housing costs by census regions (Northeast, Midwest, South, and West). It is interesting to note that the Northeast and West have higher monthly housing costs according to the figure. According to Forbes, a majority of the most overpriced cities are located in these two regions. ^3^ Figure 2 also supports this idea because median income is highest in these two regions. There is an interesting dip seen in the Northeast graph, which can be attributed to the Northeast being impacted earlier by the housing crisis.

```{r, fig.width=12, fig.height=6,echo = F, message=F,error=F, warning=FALSE, fig.cap = "Income compared to area's median grouped by house type"}
clean_years %>% 
  filter(fmtstructuretype != ".") %>% 
      # Use NSE
      group_by_("year", 
                "fmtstructuretype") %>%  
      # Take the median instead of mean to avoid extreme observations skewing the 
      # result.
      summarise_(new_y = as.formula(
        paste0("~ median(",
               "increlamipct", 
               ",na.rm=T)"))) %>%  
      # Use NSE
      ggplot(aes_string(x = "year",   
                        y = "new_y",   
                        colour = "fmtstructuretype")) +  
  geom_line(size = 1) +
  scale_y_continuous(labels = function(x){paste0(x, "%")}) +
  labs(colour = "Structure Type",
       y = "Income Percentage of Area Median Income",
       x = "Year") +
  scale_colour_brewer(palette = "Dark2") +
  theme_minimal(base_family = "serif")

```
\begin{center}
Figure 8: Income Compared to Area's Median Grouped by House Type
\end{center}

@@

```{r, fig.width=12, fig.height=6, echo = F, message=F,error=F, warning=FALSE, fig.cap="Burden based on Median Housing Cost with Assistance"}
assisted = clean_years %>% filter(!is.na(assisted)) %>% 
      # Use NSE
      group_by_("assisted", "fmtcostmedrelamicat") %>%  
      # Take the median instead of mean to avoid extreme observations skewing the 
      # result.
      summarise_(new_y= as.formula(paste0("~ median(","burden", ",na.rm=T)"))) %>% 
      ungroup %>% 
      mutate(assisted = as.factor(assisted))
ggplot(assisted, aes(y = new_y,   
                  x = fmtcostmedrelamicat,
                  fill = assisted, group = assisted)) +  
  geom_bar(
           stat="identity") + 
  labs(y="Burden",
       x = "",
       fill="Assistance") + 
  scale_y_continuous(limits=c(0,1), labels=function(x){paste0(x*100, "%")}) +
  scale_x_discrete(labels = function(x){str_replace_all(str_replace_all(x, "^[0-9] (.*)","\\1"), "LTE (.*)","<= \\1")}) +
  scale_fill_brewer("Assistance", 
                    labels = c("Without Assistance", 
                               "With Assistance"),
                    palette = "Set2")  +
  coord_flip() +
  theme_minimal(base_family = "serif")
```
\begin{center}
Figure 9: Burden based on Median housing cost with assistance
\end{center}

@@

Ultimately, our findings point towards the idea that housing affordability has decreased since the mid 80s. Although many graphs illustrate a dip in burden, this does not negate the overarching idea that housing affordability has decreased. According to Investopedia, major consequences of the mortgage crisis include chronically higher unemployment and smaller chances of getting loans and credit. The housing bubble has had effects on a long-term scale, where in 2009, there was a 38% increase in Americans filing for bankruptcy since 2007. ^4^ We conclude that the 2007 housing crisis is one of the most significant factors in causing the overall decrease in housing affordability.

@@

2. "Poverty Guidelines." ASPE. 2015. Accessed August 3, 2016. https://aspe.hhs.gov/poverty-guidelines.

3. "America's Most Overpriced Cities In 2015." Forbes. Forbes, 4 Mar. 2015. Web. 5 Aug. 2016. https://.forbes.com/sites/erincarlyle/.../americas-most-overpriced-cities-in-2015/.

4. "5 Consequences Of The Mortgage Crisis." Investopedia. N.p., 21 Nov. 2010. Web. 9 Aug. 2016. http://investopedia.com/financial-edge/1110/5-consequences-of-the-mortgage-crisis.aspx
